A User-Extensible and Adaptable Parser Architecture

### 2008-08-21 doc/sgai-2008.tex
Corrections after proofreading.

\newacronym{PDA}{Pushdown Automata}{}
\newacronym{FA}{Finite Automata}{}
\newacronym{CFL}{Context-Free Languages}{}
\newacronym{SLCT}{Simple Logfile Clustering Tool}{}
\newacronym{SMTP}{Simple Mail Transfer Protocol}{}
\newacronym{ATN}{Augmented Transition Networks}{}

@@

The architecture described herein was developed as part of a larger project
to improve anti-spam techniques by analysing the performance of the set of
filters currently in use, optimising the order and membership of the set
based on that analysis, and developing supplemental filters where there is
a deficiency.  Most anti-spam techniques are
content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms},
which requires accepting the mail in the first place, but rejecting mail
during the delivery attempt is preferable: senders of non-spam mail that is
mistakenly rejected will receive an immediate non-delivery notice; load is
reduced on the accepting mail server (allowing more intensive content-based
techniques to be used); users have less mail to wade through.  Improving
the performance of anti-spam techniques which are applied when mail is
being transferred via \SMTP{}\footnotemark{} is the goal of this project,
by providing a platform for reasoning about anti-spam filters.  The
approach chosen to measure performance is to analyse the log files produced
by the mail server, rather than modifying Postfix~\cite{postfix}, the
\SMTP{} server in use, to generate performance statistics: this approach
improves the chances of other sites testing and utilising the software.
The need arose for a parser capable of dealing with the great number and
variety of log lines produced by Postfix; the parser must be designed so
that adding support for parsing new inputs is a simple task, as the log
lines to be parsed will change over time.  The variety in log lines occurs
for several reasons:\footnotetext{Simple Mail Transfer Protocol transfers
mail across the Internet from the sender to one or more recipients.  It is
a simple, human readable, plain text protocol, making it quite easy to test
and debug problems with it.  The original protocol definition is
RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}

@@

It was hoped to reuse an existing parser rather than starting from scratch:
several parsers were considered and rejected because their parsing coverage
was too low, their parsing too inexact, and/or they did not extract
sufficient data.  The effort required to adapt and improve an existing
parser was judged to be greater than the effort to write one, as the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines with
great accuracy, but without any provision for readily parsing new or
similar log lines; others sloppily parsed the majority of log lines, but
were incapable of distinguishing between log lines of the same type, e.g.\
rejections, and extracted only a small portion of the data required.  The
only prior published work on the subject of parsing Postfix log files that
we are aware of is~\cite{log-mail-analyser}, which aims to present
correlated log data in a form suitable for a systems administrator to
search using the myriad of standard Unix text processing utilities already
available.  A full state of the art review is outside the scope of this
paper but will be included in the thesis resulting from this work.
The relationship to another procedural framework (ATN) is discussed below.

@@

    \item [Actions]  Each action performs the work required to deal with a
        category of log lines, e.g.\ processing data from rejections.

    \item [Rules]  The rules are responsible for classifying or matching
        log lines, specifying the action to invoke and the regex which
        matches the log line and extracts data.

@@

For each input log line the framework tries each rule in turn until it
finds one which matches the input log line, then invokes the action
specified by that rule.

Decoupling the parsing rules from their associated actions and the
framework allows new rules to be written and tested without requiring
modifications to the parser source code, significantly lowering the barrier
to entry for casual users who need to parse new log lines, e.g.\ part-time
systems administrators attempting to combat and reduce spam.  Decoupling
the actions from the framework simplifies both framework and actions: the
framework provides services to the actions, but does not need to perform
any tasks specific to the input being parsed; actions benefit from having
services provided by the framework, freeing them to concentrate on the task
of accurately and correctly processing the information provided by rules.

Decoupling also creates a clear separation of functionality: rules handle
low level details of identifying log lines and extracting data; actions
handle the higher level details of assembling the required data, dealing
with the intricacies of the input being parsed, complications arising,
etc.;\ the framework provides services to actions and manages the parsing
process.

There is some similarity (shown in \tableref{Similarities with ATN})
between the architecture and William Wood's 
\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
\ATN{} and the architecture is accidental, but it is interesting how two
different approaches share a similar division of responsibilities, while
having different quite semantics.

@@

\begin{table}[hbtp]
    \caption{Similarities with ATN}\label{Similarities with ATN}
    % The \smallskip{} below stop the text hitting the lines.
    \begin{tabular}[]{lll}
        \tabletopline{}%
        ATN           & Architecture    & Similarity                                \\
        \tablemiddleline{}%
        Networks      & Framework       & Determines the sequence of transitions or \\
                      &                 & actions which constitutes a valid input.  \\
        Transitions   & Actions         & Save data and impose conditions the       \\
                      &                 & input must meet to be considered valid.   \\
        Abbreviations & Rules           & Responsible for classifying input.        \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

The framework takes care of miscellaneous support functions and low level
details of parsing, freeing the authors of actions to concentrate on
writing productive code.  It provides the link between actions and rules,
allowing either to be improved independently of the other.  It provides
shared storage to pass data between actions, loads and validates rules,
controls parsing of log files, invokes actions, tracks how often each rule
matches to optimise rule ordering (\sectionref{Rule ordering}), and stores
results in the database.  The framework is largely independent of the
parser being written: the same functionality will be required for most or
all parsers, though some specialised support functions may be provided.
The framework is the core of the architecture and is deliberately quite
simple: the rules deal with the variation in log lines, and the actions
deal with the intricacies and complications encountered when parsing.

The function which finds the rule matching the log line and invokes the
requested action can be expressed in pseudo-code as:

@@

Each action contains the code required to deal with a particular type of
log line, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the required actions from scratch, unless
extending an existing parser.  In the Postfix log parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very uneven distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  It is expected that parsers
utilising this architecture will have a high ratio of rules to actions,
hopefully resulting in simpler rules and clearer distinctions between the
inputs parsed by different rules.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
action which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_DATA}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines, but that do not by themselves represent interesting events.  The
third most common action is, perhaps surprisingly, \texttt{UNINTERESTING}:
this action does nothing when executed, allowing uninteresting log lines to
be parsed without causing any effects (it does not imply that the input is
ungrammatical or unparsed).  Generally rules specifying the
\texttt{UNINTERESTING} action are rules to parse log lines which are not
associated with a specific mail, e.g.\ notices about configuration files
changing.  The framework warns about every unparsed log line, to alert the
user that they need to alter or extend their ruleset; the Postfix log
parser successfully parses every input line in the \numberOFlogFILESall{}
log files it is currently being tested with.  The remaining actions only
have one or two associated rules: in some cases this is because there will
only ever be one type of log line for that action, e.g.\ all log lines
showing that a remote client has connected are matched by a single rule and
handled by the \texttt{CONNECT} action; in other cases the action is
required to address a deficiency in the log files, or a complication which
arises during parsing.  Taking the \texttt{CONNECT} action as an example,
it creates a new data structure for each client connecting, and saves the
data extracted by the rule in memory, subsequently to be entered into the
database.  If a data structure already exists for that connection it is
treated as a symptom of a bug, and the action issues a warning containing
the full contents of the existing data structure, plus the log line which
has just been parsed.

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture: instead of writing one monolithic
function which must be modified to support new behaviour, with all the
attendant risks of adversely affecting the existing parser, when a new
requirement arises an independent action can be written to satisfy it.  In
some cases the new action will require the cooperation of other actions,
e.g.\ to set or check a flag.  There is a possibility of introducing
failure when modifying existing actions, but the modifications will be
smaller and occur less frequently than with a monolithic architecture, thus
failures will be less likely and will be easier to test for and diagnose.
The parser is written in an object oriented style, allowing sub-classes to
extend or override actions in addition to adding new actions; because each
action is an independent procedure, the sub-class need only modify the
action it is overriding, rather than reproducing large chunks of
functionality.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log lines, there were a large number of
complications to be overcome.  Some were the result of deficiencies in
Postfix's logging (some of which were improved on or resolved by later
versions of Postfix); others were due to the vagaries of process
scheduling, client behaviour, and administrative actions.  All were
successfully accommodated in the Postfix log parser: adding new actions was
enough to overcome several of the complications; others required
modifications to a single existing action to work around the difficulties;
the remainder were resolved by adapting existing actions to cooperate and
exchange extra data, changing their behaviour as appropriate.

Actions may also return a modified input line which will be parsed as if
read from the input stream, allowing for a simplified version of cascaded
parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
number of rules and actions to parse a single input line.

@@

Rules categorise inputs, specifying both the regex to match against each
log line and the action to invoke when the match is successful.  The
Postfix log parser stores its rules in the same SQL database the results
are stored in, removing any doubt as to which set of rules was used to
produce a set of results; other implementations are free to store the rules
in whatever fashion suits their needs.  The framework requires each rule to
have \texttt{action} and \texttt{regex} attributes; each implementation is
free to add any additional attributes it requires.  The Postfix log parser
adds several attributes to optimise rule ordering (\sectionref{Rule
ordering}), to restrict which log lines rules can be matched against, and
to describe each rule.  Data is automatically extracted and saved from the
log line based on keywords in the rule's regex, but there is also a
facility to specify extra data to be saved.

Parsing new inputs is generally achieved by creating a new rule that pairs
an existing action with a new regex.  Adding new rules is simple: define
all the required attributes and append the new rule to wherever the
existing rules are stored.  The Postfix log parser supplies a utility based
on \SLCT{}~\cite{slct-paper} to aid in producing regexes from previously
unparsed inputs.  Storing the rules separately allows other approaches to
rule management to be used, e.g.\ instead of manually adding new rules,
machine learning techniques could be used to automatically generate new
rules, based on the existing rules and input lines.  If this approach was
taken the choice of machine learning technique would be constrained by the
size of typical data sets (see \sectionref{Results}); techniques which
require the full data set when training would be impractical; Instance
Based Learning~\cite{instance-based-learning} techniques which
automatically determine which inputs from the training set are valuable and
which inputs can be discarded would probably reduce the data required to a
manageable size.  This avenue of research and development has not been
pursued by the authors, but could easily be undertaken independently.

The architecture does not try to detect overlapping rules: that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the first matching rule wins, and the order in which rules are tried
against each line may change between log files.  Overlapping rules are
frequently a requirement, allowing a more specific rule to match some log
lines and a more general rule to match the majority, e.g.\ separating
\SMTP{} delivery to specific sites from \SMTP{} delivery to the rest of the
world.  This can simplify both the general rule and the more specific rule;
additionally rules from different sources can be combined with a minimum of
prior cooperation or modification required.  Overlapping rules should have
a priority attribute to specify their relative ordering; negative
priorities may be useful for catchall rules.

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \FA{}; it is possible to convert between regexes and
\FA{}, so regex overlap could be detected by finding a non-empty
intersection of two \FA{}\@.  One formula for \FA{} intersection is given
in for example~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
\overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
more powerful than traditional regexes, e.g.\ it is possible to match
correctly balanced brackets nested to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to \PDA{}, so overlap detection may
require calculating the intersection of two \PDA{}\@.  \CFL{} are
equivalent in computational power to \PDA{}, and intersection of two
\CFL{s} is not closed, i.e.\ intersection can produce a language which
cannot be parsed by a \CFL{}, e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
\rightarrow a^{n}b^{n}c^{n}$, so the results of \PDA{} intersection may be
intractable in some cases.  Detecting overlap between $n$ regexes requires
$n(n-1)/2$ intersections to be calculated, resulting in $O(n^2x)$
complexity, where $O(x)$ is the complexity of calculating intersection.
This is certainly not a task to be performed on every parsing run:
detecting overlap amongst the Postfix log parser's \numberOFrules{} rules
would require calculating \numberOFruleINTERSECTIONS{} intersections.

@@

% Do not reformat this!
\begin{table}[htbp]
    \caption{Example rule}
    \empty{}\label{Example rule}
    \begin{tabular}[]{ll}
        \tabletopline{}%
        Attribute                 & Value                                            \\
        \tablemiddleline{}%
        %program                   & \daemon{smtpd}                                   \\
        regex                     & \verb!^connect from (__HOSTNAME__)\[(__IP__)\]$! \\
        action                    & CONNECT                                          \\
        %priority                  & 0                                                \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

    \item [Matching rules against log lines is simple:]  The first matching
        rule determines the action that will be invoked: there is no
        backtracking to try alternate rules, no attempt is made to pick a
        \textit{best\/} rule.

@@

    \item [Line oriented:]  The architecture is line oriented at present:
        there is no facility for rules to consume more input or push unused
        input back onto the input stream.  This was not a deliberate
        decision, rather a consequence of the line oriented nature of the
        log files; more flexible approaches could be pursued.

@@

Parsing efficiency is an obvious concern when the Postfix log parser
routinely needs to parse very large log files.  The mail server which
generated the log files used in testing the Postfix log parser accepts
approximately 10,000 mails for 700 users a day.  Median log file size is
50MB, containing 285,000 log lines --- large scale mail servers would have
much larger log files.  When generating the timing data used in this
section, \numberOFlogFILES{} log files (totaling 10.08 GB,
\numberOFlogLINEShuman{} log lines) were each parsed 10 times and the
parsing times averaged.  Saving results to the database was disabled for
the test runs, as the tests are aimed at measuring the speed of the Postfix
log parser rather than the speed of the database.  The computer used for
test runs was a Dell Optiplex 745 described in \tableref{Computer used to
generate statistics}; it was dedicated to the task of gathering statistics
from test runs, and was not used for any other purpose while test runs were
ongoing --- any services not necessary for running the tests were disabled.
Parsing all \numberOFlogFILES{} log files in one run took
\input{build/include-timing-run-duration}.

@@

An important property of a parser is how parsing time scales relative to
input size: linearly, polynomially, or exponentially?  \Graphref{Parsing
time, file size, and number of lines} shows the parsing time in seconds,
file size in MB, and number of log lines in tens of thousands, for each of
the \numberOFlogFILES{} log files.  All three lines run roughly in
parallel, giving the impression that the algorithm scales linearly with
input size.  This impression is borne out by \graphref{parsing time vs file
size vs number lines factor}: the ratios are tightly banded across the
graph, showing that the algorithm scales linearly.  The ratios increase
(i.e.\ improve) for log files 22 and 62--68 despite their large size due to
a mail forwarding loop resulting in a greatly increased number of mails
delivered and log lines generated.

\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
and number of lines}{Parsing time, file size, and number of lines}
\showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of number of
lines and file size to parsing time}{parsing time vs file size vs number
lines factor}

@@

How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that typical parsers will have a large number of rules.  There are
\numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
minimum number of rules required to parse the \numberOFlogFILES{} log files
is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
ruleset.  A second set of statistics was generated using the minimum
ruleset and the parsing times compared to those generated using the full
ruleset: the percentage parsing time increase when using the full ruleset
instead of the minimal ruleset for optimal, shuffled and reversed orderings
is shown in \graphref{Percentage parsing time increase of maximum ruleset
over minimum ruleset}.  It is clear that the increased number of rules has
a noticeable performance impact with reverse ordering, and a lesser impact
with shuffled ordering.  The optimal ordering shows a mean increase of
\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
results show that the architecture scales very well as the number of rules
increases, and that sorting the rules is an important optimisation enabling
this.  \showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage
parsing time increase of maximum ruleset over minimum ruleset}{Percentage
parsing time increase of maximum ruleset over minimum ruleset}

@@

The Postfix log parser requires two different types of coverage to be
measured: log lines parsed and mail delivery attempts correctly understood
(the former is a requirement for the latter to be achieved).  Improving the
former is less intrusive, as it just requires new rules to be written;
improving the latter is more intrusive as it requires changes to actions,
and it can also be much harder to notice a deficiency.

>>>>

\newacronym{PDA}{Pushdown Automata}{}
\newacronym{FA}{Finite Automata}{}
\newacronym{CFL}{Context-Free Languages}{}
\newacronym{SMTP}{Simple Mail Transfer Protocol}{}
\newacronym{ATN}{Augmented Transition Networks}{}

@@

The architecture described herein was developed as part of a larger project
to improve anti-spam techniques by analysing the performance of the set of
filters currently in use, optimising the order and membership of the set
based on that analysis, and developing supplemental filters where
deficiencies are identified.  Most anti-spam techniques are
content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}
which requires accepting the mail in the first place, but rejecting mail
during the delivery attempt is preferable: senders of non-spam mail that is
mistakenly rejected will receive an immediate non-delivery notice; load is
reduced on the accepting mail server (allowing more intensive content-based
techniques to be used); users have less mail to wade through.  Improving
the performance of anti-spam techniques which are applied when mail is
being transferred via \SMTP{}\footnotemark{} is the goal of this project,
by providing a platform for reasoning about anti-spam filters.  The
approach chosen to measure performance is to analyse the log files produced
by the \SMTP{} server in use, Postfix~\cite{postfix}, rather than modifying
it to generate statistics: this approach improves the chances of other
sites testing and utilising the software.  The need arose for a parser
capable of dealing with the great number and variety of log lines produced
by Postfix: the parser must be designed so that adding support for parsing
new inputs is a simple task, as the log lines to be parsed will change over
time.  The variety in log lines occurs for several
reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail across
the Internet from the sender to one or more recipients.  It is a simple,
human readable, plain text protocol, making it quite easy to test and debug
problems with it.  The original protocol definition is
RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}

@@

It was hoped to reuse an existing parser rather than starting from scratch:
several parsers were considered and rejected because: they parsed too small
a fraction of the log files; their parsing was too inexact; and/or they did
not extract sufficient data.  The effort required to adapt and improve an
existing parser was judged to be greater than the effort to write one, as
the techniques used by the existing parsers severely limited their
potential: some ignored the majority of log lines, parsing specific log
lines with great accuracy, but without any provision for readily parsing
new or similar log lines; others sloppily parsed the majority of log lines,
but were incapable of distinguishing between log lines of the same type,
e.g.\ rejections, and extracted only a small portion of the data required.
The only prior published work on the subject of parsing Postfix log files
that we are aware of is~\cite{log-mail-analyser}, which aims to present
correlated log data in a form suitable for a systems administrator to
search using the myriad of standard Unix text processing utilities already
available.  A full state of the art review is outside the scope of this
paper but will be included in the thesis resulting from this work; the
relationship to another procedural framework (ATN) is discussed below.

@@

    \item [Actions]  Each action performs the work required to deal with a
        category of inputs, e.g.\ processing data from rejections.

    \item [Rules]  The rules are responsible for classifying or matching
        inputs, specifying the action to invoke and the regex which matches
        the inputs and extracts data.

@@

For each input the framework tries each rule in turn until it finds one
which matches the input, then invokes the action specified by that rule.

Decoupling the parsing rules from their associated actions and the
framework allows new rules to be written and tested without requiring
modifications to the parser source code, significantly lowering the barrier
to entry for casual users who need to parse new inputs, e.g.\ part-time
systems administrators attempting to combat and reduce spam.  Decoupling
the actions from the framework simplifies both framework and actions: the
framework provides services to the actions, but does not need to perform
any tasks specific to the input being parsed; actions benefit from having
services provided by the framework, freeing them to concentrate on the task
of accurately and correctly processing the information provided by rules.

Decoupling also creates a clear separation of functionality: rules handle
low level details of identifying inputs and extracting data; actions handle
the higher level details of assembling the required data, dealing with the
intricacies of the input being parsed, complications arising, etc.;\ the
framework provides services to actions and manages the parsing process.

There is some similarity (shown in \tableref{Similarities with ATN})
between the architecture and William Wood's \ATN{}~\cite{atns, nlpip}, a
tool used in Computational Linguistics for creating grammars to parse or
generate sentences.  The resemblance between \ATN{} and the architecture is
accidental, but it is interesting how two different approaches share a
similar division of responsibilities, while having different semantics.

@@

\begin{table}[hbtp]
    \caption{Similarities with ATN}\label{Similarities with ATN}
    % The \smallskip{} below stop the text hitting the lines.
    \begin{tabular}[]{lll}
        \tabletopline{}%
        ATN           & Architecture    & Similarity                                \\
        \tablemiddleline{}%
        Networks      & Framework       & Determines the sequence of transitions or \\
                      &                 & actions which constitutes a valid input.  \\
        Transitions   & Actions         & Saves data and imposes conditions the     \\
                      &                 & input must meet to be considered valid.   \\
        Abbreviations & Rules           & Responsible for classifying input.        \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

The framework takes care of miscellaneous support functions and low level
details of parsing, freeing the authors of actions to concentrate on
writing productive code.  It provides the link between actions and rules,
allowing either to be improved independently of the other.  It provides
shared storage to pass data between actions, loads and validates rules,
manages parsing, invokes actions, tracks how often each rule matches to
optimise rule ordering (\sectionref{Rule ordering}), and stores results in
the database.  The framework is largely independent of the parser being
written: the same functionality will be required for most or all parsers,
though some specialised support functions may be provided.  The framework
is the core of the architecture and is deliberately quite simple: the rules
deal with the variation in inputs, and the actions deal with the
intricacies and complications encountered when parsing.

The function which finds the rule matching the input and invokes the
requested action can be expressed in pseudo-code as:

@@

Each action contains the code required to deal with a particular category
of input, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the required actions from scratch, unless
extending an existing parser.  In the Postfix log parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very uneven distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  It is expected that parsers
utilising this architecture will have a high ratio of rules to actions,
hopefully resulting in simpler rules and clearer distinctions between the
inputs parsed by different rules.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
action which handles Postfix rejecting a delivery attempt; it is followed
by \texttt{SAVE\_DATA}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines, but that do not by themselves represent interesting events.  The
third most common action is, perhaps surprisingly, \texttt{UNINTERESTING}:
this action does nothing when executed, allowing uninteresting log lines to
be parsed without causing any effects (it does not imply that the input is
ungrammatical or unparsed).  Generally rules specifying the
\texttt{UNINTERESTING} action are rules to parse log lines which are not
associated with a specific mail, e.g.\ notices about configuration files
changing.  The framework warns about every unparsed input, to alert the
user that they need to alter or extend their ruleset; the Postfix log
parser successfully parses every log line in the \numberOFlogFILESall{} log
files it is currently being tested with.  The remaining actions only have
one or two associated rules: in some cases this is because there will only
ever be one log line variant for that action, e.g.\ all log lines showing
that a remote client has connected are matched by a single rule and handled
by the \texttt{CONNECT} action; in other cases the action is required to
address a deficiency in the log files, or a complication which arises
during parsing.  Taking the \texttt{CONNECT} action as an example, it
creates a new data structure in memory for each client connecting, and
saving the data extracted by the rule; this data will subsequently be
entered into the database.  If a data structure already exists for that
connection it is treated as a symptom of a bug, and the action issues a
warning containing the full contents of the existing data structure, plus
the log line which has just been parsed.

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture.  Instead of writing one monolithic
function which must be modified to support new behaviour, with all the
attendant risks of adversely affecting the existing parser, when a new
requirement arises an independent action can be written to satisfy it.  In
some cases the new action will require the cooperation of other actions,
e.g.\ to set or check a flag.  There is a possibility of introducing
failure when modifying existing actions, but the modifications will be
smaller and occur less frequently than with a monolithic architecture, thus
failures will be less likely and will be easier to test for and diagnose.
The architecture can be implemented in an object oriented style, allowing
sub-classes to extend or override actions in addition to adding new
actions; because each action is an independent procedure, the sub-class
need only modify the action it is overriding, rather than reproducing large
chunks of functionality.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log lines, there were a large number of
complications to be overcome (covered in detail in the thesis).  Some were
the result of deficiencies in Postfix's logging (some of which were
improved on or resolved by later versions of Postfix); others were due to
the vagaries of process scheduling, client behaviour, and administrative
actions.  All were successfully accommodated in the Postfix log parser:
adding new actions was enough to overcome several of the complications;
others required modifications to a single existing action to work around
the difficulties; the remainder were resolved by adapting existing actions
to cooperate and exchange extra data, changing their behaviour as
appropriate.

Actions may return a modified input line which will be parsed as if read
from the input stream, allowing for a simplified version of cascaded
parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
number of rules and actions to parse a single input line.

@@

Rules categorise inputs, specifying both the regex to match against each
input and the action to invoke when the match is successful.  The Postfix
log parser stores its rules in the same SQL database the results are stored
in, removing any doubt as to which set of rules was used to produce a set
of results; other implementations are free to store the rules in whatever
fashion suits their needs.  The framework requires each rule to have
\texttt{action} and \texttt{regex} attributes; each implementation is free
to add any additional attributes it requires.  The Postfix log parser adds
several attributes to optimise rule ordering (\sectionref{Rule ordering}),
to restrict which log lines each rule can be matched against, and to
describe each rule.  Data is automatically extracted and saved from the log
line based on keywords in the rule's regex, but there is also a facility to
specify extra data to be saved.

Parsing new inputs is generally achieved by creating a new rule that pairs
an existing action with a new regex.  Adding new rules is simple: define
all the required attributes and append the new rule to wherever the
existing rules are stored.  The Postfix log parser supplies a utility based
on Simple Logfile Clustering Tool~\cite{slct-paper} to aid in producing
regexes from previously unparsed log lines.  Other approaches to rule
management can be used, e.g.\ instead of manually adding new rules, machine
learning techniques could be used to automatically generate new rules,
based on the existing rules and input lines.  If this approach was taken
the choice of machine learning technique would be constrained by the size
of typical data sets (see \sectionref{Results}).  Techniques which require
the full data set when training would be impractical: Instance Based
Learning~\cite{instance-based-learning} techniques which automatically
determine which inputs from the training set are valuable and which inputs
can be discarded would probably reduce the data required to a manageable
size.  This avenue of research and development has not been pursued by the
authors, but could easily be undertaken independently.

The architecture does not try to detect overlapping rules: that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the first matching rule wins, and the order in which rules are tried
against each input may change between parser invocations.  Overlapping
rules are frequently a requirement, allowing a more specific rule to match
some inputs and a more general rule to match the majority, e.g.\ separating
\SMTP{} delivery to specific sites from \SMTP{} delivery to the rest of the
world.  This can simplify both the general rule and the more specific rule;
additionally rules from different sources can be combined with a minimum of
prior cooperation or modification required.  Overlapping rules should have
a priority attribute to specify their relative ordering; negative
priorities may be useful for catchall rules.

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \FA{}; it is possible to convert between regexes and
\FA{}, so regex overlap could be detected by finding a non-empty
intersection of two \FA{}\@.  One formula for \FA{} intersection is given
in for example~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
\overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
more powerful than traditional regexes, e.g.\ it is possible to match
correctly balanced brackets nested to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to \PDA{}, so overlap detection may
require calculating the intersection of two \PDA{}\@.  The results of
\PDA{} intersection may be intractable in some cases: the intersection of
two \CFL{} (equivalent in computational power to \PDA{}) is not closed,
i.e.\ the resulting language cannot in all cases be parsed by a \CFL{}\@.
A simple example is:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.

Detecting overlap between $n$ regexes requires $n(n-1)/2$
intersections to be calculated, resulting in $O(n^2x)$ complexity, where
$O(x)$ is the complexity of calculating intersection.  This is certainly
not a task to be performed on every parsing run: detecting overlap amongst
the Postfix log parser's \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

@@

% Do not reformat this!
\begin{table}[htbp]
    \caption{Example rule}
    \empty{}\label{Example rule}
    \begin{tabular}[]{ll}
        \tabletopline{}%
        Attribute                 & Value                                            \\
        \tablemiddleline{}%
        %program                   & \daemon{smtpd}                                   \\
        regex                     & \verb!^connect from (__CLIENT_HOSTNAME__)\[(__CLIENT_IP__)\]$! \\
        action                    & CONNECT                                          \\
        %priority                  & 0                                                \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

    \item [Matching rules against inputs is simple:]  The first matching
        rule determines the action that will be invoked: there is no
        backtracking to try alternate rules, no attempt is made to pick a
        \textit{best\/} rule.

@@

    \item [Line oriented:]  The architecture is line oriented at present:
        there is no facility for rules to consume more input or push unused
        input back onto the input stream.  This was not a deliberate
        decision, rather a consequence of the line oriented nature of the
        Postfix log files; more flexible approaches could be pursued.

@@

Parsing efficiency is an obvious concern when the Postfix log parser
routinely needs to parse large log files.  The mail server which generated
the log files used in testing the Postfix log parser accepts approximately
10,000 mails for 700 users per day.  Median log file size is 50MB,
containing 285,000 log lines --- large scale mail servers would have much
larger log files.  When generating the timing data used in this section,
\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
log lines) were each parsed 10 times and the parsing times averaged.
Saving results to the database was disabled for the test runs, as the tests
are aimed at measuring the speed of the Postfix log parser rather than the
speed of the database.  The computer used for test runs was a Dell Optiplex
745 described in \tableref{Computer used to generate statistics}; it was
dedicated to the task of gathering statistics from test runs, and was not
used for any other purpose while test runs were ongoing --- any services
not necessary for running the tests were disabled.  Parsing all
\numberOFlogFILES{} log files in one run took
\input{build/include-timing-run-duration}.

@@

An important property of a parser is how parsing time scales relative to
input size: linearly, polynomially, or exponentially?  \Graphref{Parsing
time, file size, and number of lines} shows the parsing time in seconds,
file size in MB, and number of log lines in tens of thousands, for each of
the \numberOFlogFILES{} log files.  All three lines run roughly in
parallel, giving the impression that the algorithm scales linearly with
input size.  This impression is borne out by \graphref{parsing time vs file
size vs number lines factor}: the ratios are tightly banded across the
graph, showing that the algorithm scales linearly.  The ratios increase
(i.e.\ improve) for log files 22 and 62--68 despite their large size; that
unusually large size is due to a mail forwarding loop resulting in a
greatly increased number of mails delivered and log lines generated.

\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
and number of lines}{Parsing time, file size, and number of lines}
\showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of number of
log lines and file size to parsing time}{parsing time vs file size vs
number lines factor}

@@

How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that typical parsers will have a large number of rules.  There are
\numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
minimum number of rules required to parse the \numberOFlogFILES{} log files
is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
ruleset.  A second set of statistics was generated using the minimum
ruleset, and the parsing times compared to those generated using the full
ruleset: the percentage parsing time increase when using the full ruleset
instead of the minimal ruleset for optimal, shuffled and reversed orderings
is shown in \graphref{Percentage parsing time increase of maximum ruleset
over minimum ruleset}.  It is clear that the increased number of rules has
a noticeable performance impact with reverse ordering, and a lesser impact
with shuffled ordering.  The optimal ordering shows a mean increase of
\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
results show that the architecture scales very well as the number of rules
increases, and that optimally ordering the rules is an important
optimisation enabling this.
\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}

@@

The Postfix log parser has two different types of coverage to be measured:
log lines correctly parsed, and mail delivery attempts correctly understood
(the former is a requirement for the latter to be achieved).  Improving the
former is less intrusive, as usually it just requires new rules to be
written; improving the latter is more intrusive as it requires changes to
actions, and it can also be much harder to notice that a deficiency exists.
